Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-02 article.tex

\subsection{Ordinary differential equation models}
Dynamic models describe systems with states $x$, usually quantifying the involved species, interacting and evolving over time. %  as e.g. given by chemical reaction networks. 
The time evolution of the states is expressed via a set of ODEs, $\dot x = f(x)$. Although constituting quite a special class of dynamic systems, chemical reaction networks formulated by the law of mass action allow for surprisingly general applications. Typical ODE examples derived from the law of mass action are:
\begin{itemize}
	\item $\dot x_A = k_p$ constant production of $x_A$ ($\emptyset \rightarrow A$)
	\item $\dot x_C = k_c \cdot x_A \cdot x_B = - \dot x_A = - \dot x_B$ complex formation ($A+B \rightarrow C$)
	\item $\dot x_C = - k_d \cdot x_C$ constant degradation of $x_C$ ($C \rightarrow \emptyset$)
\end{itemize}
each generally written as 
\begin{align}
	\dot x = \underbrace{k_0 \left(\prod x_i^{\nu_i}\right)}_{v_0(x, k_0)} (\nu' - \nu) \quad\leftrightarrow\quad (\nu_1 X_1 + \dots + \nu_n X_n \longrightarrow \nu'_1 X_1 + \dots + \nu'_n X_n)
	\label{eq:ma}
\end{align}
where $X_1, \dots, X_n$ are the $n$ species involved, $\nu$ and $\nu'$ encode the stoichiometry of the reaction and $k_0$ is the reaction rate. When several reactions are involved, eq.~\eqref{eq:ma} generalizes to 
\begin{align}
	\dot x = S\cdot v(x, k)
	\label{eq:Sv}
\end{align}
%$\dot x_l = \sum_{j = 1\dots m}  k_{j} \prod_{i = 1\dots n} \nu_{i,j} x_i^{\nu_{i,j}}$ for a system of $n$ states and $m$ reactions with rate constants $p = \{ k_j, j = 1\dots m\}$ and stoichiometric constants $\nu_{i,j} \in \mathbb{Z}$. 
%An even more general expression of the system is given by $\dot x = S v(x,p)$ 
with the stoichiometry matrix $S$
%containing the stoichiometric constants $\nu_{i,j}$ 
obtained from the collection of vectors $\nu' - \nu$ for reactions $1, \dots, r$,
and the flux vector $v(x,k)\in\mathbb R^{r}$.
Eq.~\eqref{eq:Sv} also holds for general ODEs that do not follow from mass action kinetics. Moreover the system may be explicitly time-dependent and may contain forcings $u(t)$ which e.g.~describe the external stimulation of the system, $\dot x = S \cdot v(x,k,u,t)$. %Last but not least, the solution of eq.~\eqref{eq:Sv} depends on initial values $x_0 = x(t = 0)$.

Often the experimentally observed quantities $y$ do not directly correspond to the species $x$ described by the model, but are obtained via an observation function $g: \mathbb{R}^n \rightarrow \mathbb{R}^m$,  $y = g(x)$. Typical examples are unknown scalings and offsets in the measurement $y = c_s \cdot x + c_o$ or the joint measurement of several species $y_i = \sum_j c_j x_j$.   \\
The model prediction is therefore obtained by evaluating the observation function on the integrated ODEs, given a set of free parameters $\theta$ of the system. The parameters $\theta$ contain the rate constants, initial values $x_0 = x(t=0)$ and additional parameters included in the prediction function and forcings. It is generally assumed that the stoichiometry of the system is known. \\
Besides the model predition itself, also the sensitivity of the prediction to changes in the parameter values, $\frac{\partial y}{\partial \theta}$, is of interest. Sensitivity equations can be directly inferred

>>>>

\subsection{Dynamic models and model sensitivities}
Dynamic models describe systems with states $x$, usually quantifying the involved species, interacting and evolving over time. %  as e.g. given by chemical reaction networks. 
The time evolution of the states is expressed via a set of ODEs, $\dot x = f(x)$. Although constituting quite a special class of dynamic systems, chemical reaction networks formulated by the law of mass action allow for surprisingly general applications. Typical ODE examples derived from the law of mass action are:
\begin{itemize}
	\item $\dot x_A = k_p$ constant production of $x_A$ ($\emptyset \rightarrow A$)
	\item $\dot x_C = k_c \cdot x_A \cdot x_B = - \dot x_A = - \dot x_B$ complex formation ($A+B \rightarrow C$)
	\item $\dot x_C = - k_d \cdot x_C$ constant degradation of $x_C$ ($C \rightarrow \emptyset$)
\end{itemize}
each generally written as 
\begin{align}
	\dot x = \underbrace{k_0 \left(\prod x_i^{\nu_i}\right)}_{v_0(x, k_0)} (\nu' - \nu) \quad\leftrightarrow\quad (\nu_1 X_1 + \dots + \nu_n X_n \longrightarrow \nu'_1 X_1 + \dots + \nu'_n X_n)
	\label{eq:ma}
\end{align}
where $X_1, \dots, X_n$ are the $n$ species involved, $\nu$ and $\nu'$ encode the stoichiometry of the reaction and $k_0$ is the reaction rate. When several reactions are involved, eq.~\eqref{eq:ma} generalizes to 
\begin{align}
	\dot x = S\cdot v(x, k)
	\label{eq:Sv}
\end{align}
%$\dot x_l = \sum_{j = 1\dots m}  k_{j} \prod_{i = 1\dots n} \nu_{i,j} x_i^{\nu_{i,j}}$ for a system of $n$ states and $m$ reactions with rate constants $p = \{ k_j, j = 1\dots m\}$ and stoichiometric constants $\nu_{i,j} \in \mathbb{Z}$. 
%An even more general expression of the system is given by $\dot x = S v(x,p)$ 
with the stoichiometry matrix $S$
%containing the stoichiometric constants $\nu_{i,j}$ 
obtained from the collection of vectors $\nu' - \nu$ for reactions $1, \dots, r$,
and the flux vector $v(x,k)\in\mathbb R^{r}$.
Eq.~\eqref{eq:Sv} also holds for general ODEs that do not follow from mass action kinetics. Moreover the system may be explicitly time-dependent and may contain forcings $u(t)$ which e.g.~describe the external stimulation of the system. A general form of the dynamic model is therefore given by
\begin{align} 
	\dot x = f(x,k,u,t) = S \cdot v(x,k,u,t), \quad\quad \textrm{with }  x(t = 0) = x_0.
\end{align}
The solution $x(t,\theta)$ for a given parameter vector $\theta = (k, x_0)$ is called model prediction. %The parameters $\theta$ contain the rate constants, initial values $x_0$ and additional parameters included in the forcings. 
It is generally assumed that the stoichiometry matrix $S$ of the system is known.
Besides the model predition itself, also the sensitivity of the prediction to changes in the parameter values is of interest. 
The sensitivities $s_i = \frac{\partial x}{\partial \theta_i}$ satisfy the \textit{sensitivity equations}
\begin{align}
	\dot s_i  = \frac{\partial f}{\partial x} s_i + \frac{\partial f}{\partial \theta_i}, \qquad \textrm{with }s_i(t = 0) = \frac{\partial x_0}{\partial \theta_i},
\end{align}
a system of ordinary differential equations that in general depend on the states $x$ and, therefore, need to be solved jointly with the original ODE.

Often the experimentally observed quantities $y$ do not directly correspond to the species $x$ described by the ODE, but are obtained via an observation function $g: \mathbb{R}^n \rightarrow \mathbb{R}^m$,  \begin{align}
	y = g(x, c)
\end{align}
with the observation parameters $c$. Examples for observation functions are scaling and offset transformations, $y = c_s \cdot x + c_o$, or the measurement of a superposition of species, $y_i = \sum_j c_j x_j$. The model prediction for the observed states is obtained by evaluating the observation function on the solution of the ODE. Following the chain rule of differentation also the sensitivities of the observed states $\frac{\partial y}{\partial \theta_i} = \frac{\partial g}{\partial x} \frac{\partial x}{\partial\theta_i}  + \frac{\partial g}{\partial \theta_i}$ are obtained. Here, the parameter vector $\theta$ has been augmented by the observation parameters, $\theta = (k, x_0, c)$.

The estimation of the parameters $\theta$ given the observation $y(t)$ is addressed in the next section.
%Next, the question how to estimate the parameters $\theta$ given the observation $y(t)$ is addressed.
